### 2019-08-01 reports/thesis/chapters/3_compoundPoisson.tex
CompoundPoisson chapter
Added 3rd order moment
Use multlined in align

Returning back to the compound Poisson with no electronic noise $X|Y = \sum_{i=1}^{Y}U_i$. Let the moment generating function (m.g.f.) of $X$ be $M_X(\theta)=\expectation\left[\euler^{X\theta}\right]$. It can be shown that the m.g.f.~is
\begin{equation}
  M_X(\theta)=
  \exp\left[
    \lambda
    \left(
      M_U(\theta)-1
    \right)
  \right]
\end{equation}
\citep{gatto2010saddlepoint}. The derivation is shown in Appendix \ref{chapter:appendix_compoundPoissonMgf}. Moments of $X$ can be obtained from the m.g.f.~by differentiating it and setting it to zero. In other words, $\expectation[X^r]=M_X^{(r)}(\theta)$. Then it can be shown that
\begin{align}
  \expectation\left[X\right]&=\lambda\expectation\left[U\right]
  \\
  \variance\left[X\right] &= \lambda\expectation\left[U^2\right]
\end{align}
which agrees with the results in the previous section. The derivation is shown in Appendix \ref{chapter:appendix_compoundPoissonMgf}.

@@

The m.g.f.~of $U$ is $M_U(\theta) = \left(\dfrac{\beta}{\beta-\alpha}\right)^\alpha$. Then the m.g.f.~of $X$ is
\begin{equation}
  M_X(\theta)=\exp\left[\lambda\left(\left(\frac{\beta}{\beta-\theta}\right)^{\alpha}-1\right)\right]
\end{equation}
and moments can be obtained from it such as
\begin{equation}
  \expectation\left[X\right]=\frac{\alpha\lambda}{\beta}
\end{equation}
\begin{equation}
  \variance\left[X\right]=\frac{\alpha(\alpha+1)\lambda}{\beta^2}
  \label{eq:compoundPoisson_variance}
\end{equation}
and
\begin{equation}
  \expectation\left[(X-\expectation[X])^3\right] = \frac{\alpha(\alpha+1)(\alpha+2)\lambda}{\beta^3}
  \ .
\end{equation}

@@

Let $\widehat{\lambda}$, $\widehat{\alpha}$ and $\widehat{\beta}$ be estimators of $\lambda$, $\alpha$ and $\beta$ respectively. The log likelihood is defined to be
\begin{equation*}
  \ln L(\lambda,\alpha,\beta;X) = \sum_{i=1}^n
  \left[
    \mathbb{I}(x_i=0)\ln \prob(X=x_i)
    +\mathbb{I}(x_i>0)\ln p_X(x_i)
  \right]
\end{equation*}
\begin{multline}
  \ln L(\lambda,\alpha,\beta;X) = \sum_{i=1}^n
  \left[
    -\mathbb{I}(x_i=0)\lambda
    \vphantom{\sum_i^i}
  \right.
  \\
  \left.
    +\mathbb{I}(x_i>0)\left(
    -1-\beta x_i-\lambda-\ln x -\ln\sum_{y=1}^\infty W_y
    \right)
  \right] \ .
\end{multline}
$\widehat{\lambda}$, $\widehat{\alpha}$ and $\widehat{\beta}$ are values of $\lambda$, $\alpha$ and $\beta$ which jointly maximises the log likelihood.

@@

In the M step, the conditional expected joint log likelihood is maximised with respect to the parameters $\lambda$, $\alpha$ and $\beta$. The objective function is then
\begin{multline*}
  T(\lambda,\alpha,\beta)=
  \sum_{i=1}^n
  \expectation\left[
    -\mathbb{I}(x_i=0)
    \lambda
  \right.
  \\
  \left.+
    \mathbb{I}(x_i>0)
    \left(
      Y_i\alpha\ln\beta-\ln\Gamma(Y_i\alpha)+(Y_i\alpha-1)\ln x_i - \beta x_i
    \right.
  \right.
  \\
  \left.
    \left.  
      - \lambda + Y_i \ln \lambda - \ln(Y_i!)
    \right)
    |X_i=x_i
  \right]
\end{multline*}
\begin{multline}
  T(\lambda,\alpha,\beta)=
  -n\lambda
  \\
  +\sum_{i=1}^n
  \mathbb{I}(x_i>0)
  \left[
    \expectation[Y_i|X_i=x_i]\alpha\ln\beta-\expectation[\ln\Gamma(Y_i\alpha)|X_i=x_i]
  \right.
  \\
  \left.
    +\expectation[Y_i|X_i=x_i]\alpha\ln x_i - \beta x_i
    + \expectation[Y_i|X_i=x_i] \ln \lambda
  \right] + c
\end{multline}
where $c$ is some constant not dependent on $\lambda$, $\alpha$ or $\beta$.

@@

The Hessian matrix is, omitting the $\mathbb{I}(x_i>0)$ term,
\begin{equation}
  \nabla_{\alpha,\beta}\nabla_{\alpha,\beta}\T T \approx
  \sum_{i=1}^n
  \begin{pmatrix}
    -y_i/\alpha  & y_i/\beta \\
    y_i/\beta & -\alpha y_i/\beta^2
  \end{pmatrix}
  \ . 
\end{equation}
The determinant of the Hessian matrix is then
\begin{align*}
  \|\nabla_{\alpha,\beta}\nabla_{\alpha,\beta}\T T\|
  &\approx
  \left(\sum_{i=1}^n-\frac{\alpha y_i}{\beta^2}\right)
  \left(\sum_{i=1}^n-\frac{y_i}{\alpha}\right) - \left(\sum_{i=1}^n\frac{y_i}{\beta}\right)^2
  \\
  \|\nabla_{\alpha,\beta}\nabla_{\alpha,\beta}\T T\|
  &\approx
  \left(\sum_{i=1}^n\sum_{j=1}^n\frac{y_iy_j}{\beta^2}\right)
   - \left(\sum_{i=1}^n\frac{y_i}{\beta}\right)\left(\sum_{j=1}^n\frac{y_j}{\beta}\right)
  \\
  \|\nabla_{\alpha,\beta}\nabla_{\alpha,\beta}\T T\|
  &\approx
  \left(\sum_{i=1}^n\sum_{j=1}^n\frac{y_iy_j}{\beta^2}\right)
   - \left(\sum_{i=1}^n\sum_{j=1}^n\frac{y_iy_j}{\beta^2}\right)
\end{align*}
to get
\begin{equation}
  \|\nabla_{\alpha,\beta}\nabla_{\alpha,\beta}\T T\|
  \approx
  0
  \ .
\end{equation}

@@

The parameter $\beta$ can be constrained for a given mean to be
\begin{equation}
  \beta = \frac{\lambda\alpha}{\widehat{\mu}}
  \label{eq:compoundPoisson:beta_restrict}
\end{equation}
where
\begin{equation}
  \widehat{\mu} = \frac{1}{n}\sum_{i=1}^n x_i
\end{equation}
and it was investigated whenever a unique solution to $\dfrac{\partial T}{\partial \alpha} = 0$ can be found for high $\lambda$ and $\alpha$. The constrained objective function is
\begin{multline}
  T(\lambda,\alpha)
  =
  -n\lambda+
  \sum_{i=1}^n
  \mathbb{I}(x_i>0)
  \left[
  y_i\alpha
  \ln\left(\frac{\lambda\alpha}{\widehat{\mu}}\right)-\ln\Gamma(\alpha y_i) - \frac{1}{2}\zeta_i\alpha^2\psi'(y_i\alpha)
  \right.
  \\
  \left.
    +y_i\alpha\ln x_i - \frac{\lambda\alpha x_i}{\widehat{\mu}}
    + y_i \ln \lambda
    \vphantom{\frac{1}{1}}
  \right]
  + c
\end{multline}
which can be approximated to
\begin{multline*}
  T(\lambda,\alpha)=
  -n\lambda+
  \sum_{i=1}^n
  \mathbb{I}(x_i>0)
  \left[
    y_i\alpha\ln\left(\frac{\lambda\alpha}{\widehat{\mu}}\right)-\alpha y_i\ln(\alpha y_i) + \alpha y_i - \frac{1}{2}\frac{\zeta_i\alpha}{y_i}
  \right.
  \\
  \left.
    +y_i\alpha\ln x_i - \frac{\lambda\alpha x_i}{\widehat{\mu}}
    + y_i \ln \lambda
    \vphantom{\frac{1}{1}}
  \right]
  + c
\end{multline*}
\begin{multline}
  T(\lambda,\alpha)=
  -n\lambda+
  \sum_{i=1}^n
  \mathbb{I}(x_i>0)
  \left[
    y_i\alpha\ln\left(\frac{\lambda}{y_i\widehat{\mu}}\right) + \alpha y_i - \frac{1}{2}\frac{\zeta_i\alpha}{y_i}
  \right.
  \\
  \left.
    +y_i\alpha\ln x_i - \frac{\lambda\alpha x_i}{\widehat{\mu}}
    + y_i \ln \lambda
    \vphantom{\frac{1}{1}}
  \right]
  + c
  \ .
\end{multline}
Taking the derivative with respect to $\alpha$
\begin{equation}
  \frac{\partial T}{\partial \alpha} =
  \sum_{i=1}^n
  \mathbb{I}(x_i>0)
  \left[
    y_i\ln\left(\frac{\lambda}{y_i\widehat{\mu}}\right)
    + y_i - \frac{1}{2}\frac{\zeta_i}{y_i}
    +y_i\ln x_i - \frac{\lambda x_i}{\widehat{\mu}}
    \vphantom{\frac{1}{1}}
  \right]
\end{equation}
and this is not $\alpha$ dependent, thus there is no solution to $\alpha$.

@@

The log likelihood, using a simulation of 100 compound Poisson-gamma random variables, are shown in Figure \ref{fig:compoundPoisson_loglikelihoodplot}. These plots are concerning because the likelihood appeared not to be convex and there is no unique maxima, surprising in $\lambda=1$ case as well.

@@

These plots suggest there is no unique maximum likelihood estimator and the EM algorithm would be very sensitive to the starting position, including for small $\lambda$.

@@

XCT is in the realm of high $\lambda$ and this can be shown using back of the envelope calculations. Using the properties in the \texttt{AbsNoFilter} dataset, the power and potential difference of the x-ray tube are $P=\SI{36}{\watt}$ and $V=\SI{80}{\kilo\volt}$ respectively. The time exposure is $\tau=\SI{708}{\milli\second}$. Assume an efficiency of \SI{0.01}{{photon}\,{electron}^{-1}} and the charge per electron is $e=\SI{1.602e-19}{\coulomb\,{electron}^{-1}}$. Assume the photons projected onto the array of $2\,000^2\,\si{\pixel}$. The number of photons detected per pixel is
\begin{align*}
\lambda &\approx \dfrac{\text{efficiency} P \tau}{\text{number of pixels} V e}
\\
&\sim 10^7\,\si{{photon}\,\pixel^{-1}}
\ .
\end{align*}
Fitting the compound Poisson model onto XCT data would not be possible with such a high value of $\lambda$.

>>>>

Returning back to the compound Poisson with no electronic noise $X|Y = \sum_{i=1}^{Y}U_i$. Let the moment generating function (m.g.f.) of $X$ be $M_X(\theta)=\expectation\left[\euler^{X\theta}\right]$. It can be shown that the m.g.f.~is
\begin{equation}
  M_X(\theta)=
  \exp\left[
    \lambda
    \left(
      M_U(\theta)-1
    \right)
  \right]
\end{equation}
\citep{gatto2010saddlepoint}. The derivation is shown in Appendix \ref{chapter:appendix_compoundPoissonMgf}. Moments of $X$ can be obtained from the m.g.f.~by differentiating it and setting it to zero. In other words, $\expectation[X^r]=M_X^{(r)}(\theta)$. Then it can be shown that
\begin{align}
  \expectation\left[X\right]&=\lambda\expectation\left[U\right]
  \\
  \variance\left[X\right] &= \lambda\expectation\left[U^2\right]
  \\
  \expectation\left[(X-\expectation[X])^3\right] &=\lambda\expectation\left[U^3\right]
\end{align}
which agrees with the expectation and variance results in the previous section. The derivation is shown in Appendix \ref{chapter:appendix_compoundPoissonMgf}.

@@

The m.g.f.~of $U$ is $M_U(\theta) = \left(\dfrac{\beta}{\beta-\theta}\right)^\alpha$. Then the m.g.f.~of $X$ is
\begin{equation}
  M_X(\theta)=\exp\left[\lambda\left(\left(\frac{\beta}{\beta-\theta}\right)^{\alpha}-1\right)\right]
\end{equation}
and moments can be obtained from it such as
\begin{equation}
  \expectation\left[X\right]=\frac{\alpha\lambda}{\beta}
\end{equation}
\begin{equation}
  \variance\left[X\right]=\frac{\alpha(\alpha+1)\lambda}{\beta^2}
  \label{eq:compoundPoisson_variance}
\end{equation}
and
\begin{equation}
  \expectation\left[(X-\expectation[X])^3\right] = \frac{\alpha(\alpha+1)(\alpha+2)\lambda}{\beta^3}
  \ .
\end{equation}

@@

Let $\widehat{\lambda}$, $\widehat{\alpha}$ and $\widehat{\beta}$ be estimators of $\lambda$, $\alpha$ and $\beta$ respectively. The log likelihood is defined to be
\begin{align}
  \ln L(\lambda,\alpha,\beta;X) &= \sum_{i=1}^n
  \left[
    \mathbb{I}(x_i=0)\ln \prob(X=x_i)
    +\mathbb{I}(x_i>0)\ln p_X(x_i)
  \right]
  \nonumber \\
  &\!\begin{multlined}[b]
  =
    \sum_{i=1}^n
    \left[
      -\mathbb{I}(x_i=0)\lambda
      \vphantom{\sum_i^i}
    \right.
    {}\\
    \left.
      +\mathbb{I}(x_i>0)\left(
      -1-\beta x_i-\lambda-\ln x -\ln\sum_{y=1}^\infty W_y
      \right)
    \right] \ .
  \end{multlined}
\end{align}

$\widehat{\lambda}$, $\widehat{\alpha}$ and $\widehat{\beta}$ are values of $\lambda$, $\alpha$ and $\beta$ which jointly maximises the log likelihood.

@@

In the M step, the conditional expected joint log likelihood is maximised with respect to the parameters $\lambda$, $\alpha$ and $\beta$. The objective function is then
\begin{align}
  T(\lambda,\alpha,\beta)&=
  \begin{multlined}[t]
    \sum_{i=1}^n
    \expectation\left[
      -\mathbb{I}(x_i=0)
      \lambda
    \right.
    \\
    \left.+
      \mathbb{I}(x_i>0)
      \left(
        Y_i\alpha\ln\beta-\ln\Gamma(Y_i\alpha)+(Y_i\alpha-1)\ln x_i - \beta x_i
      \right.
    \right.
    \\
    \left.
      \left.  
        - \lambda + Y_i \ln \lambda - \ln(Y_i!)
      \right)
      |X_i=x_i
    \right]
  \end{multlined}
  \nonumber\\
  &\!\begin{multlined}[b]=
    -n\lambda
    \\
    +\sum_{i=1}^n
    \mathbb{I}(x_i>0)
    \left[
      \expectation[Y_i|X_i=x_i]\alpha\ln\beta-\expectation[\ln\Gamma(Y_i\alpha)|X_i=x_i]
    \right.
    {}\\
    \left.
      +\expectation[Y_i|X_i=x_i]\alpha\ln x_i - \beta x_i
      + \expectation[Y_i|X_i=x_i] \ln \lambda
    \right] + c
  \end{multlined}
\end{align}
where $c$ is some constant not dependent on $\lambda$, $\alpha$ or $\beta$.

@@

The Hessian matrix is, omitting the $\mathbb{I}(x_i>0)$ term,
\begin{equation}
  \nabla_{\alpha,\beta}\nabla_{\alpha,\beta}\T T \approx
  \sum_{i=1}^n
  \begin{pmatrix}
    -y_i/\alpha  & y_i/\beta \\
    y_i/\beta & -\alpha y_i/\beta^2
  \end{pmatrix}
  \ . 
\end{equation}
The determinant of the Hessian matrix is then
\begin{align*}
  \|\nabla_{\alpha,\beta}\nabla_{\alpha,\beta}\T T\|
  &\approx
  \left(\sum_{i=1}^n-\frac{\alpha y_i}{\beta^2}\right)
  \left(\sum_{i=1}^n-\frac{y_i}{\alpha}\right) - \left(\sum_{i=1}^n\frac{y_i}{\beta}\right)^2
  \\
  &\approx
  \left(\sum_{i=1}^n\sum_{j=1}^n\frac{y_iy_j}{\beta^2}\right)
   - \left(\sum_{i=1}^n\frac{y_i}{\beta}\right)\left(\sum_{j=1}^n\frac{y_j}{\beta}\right)
  \\
  &\approx
  \left(\sum_{i=1}^n\sum_{j=1}^n\frac{y_iy_j}{\beta^2}\right)
   - \left(\sum_{i=1}^n\sum_{j=1}^n\frac{y_iy_j}{\beta^2}\right)
\end{align*}
to get
\begin{equation}
  \|\nabla_{\alpha,\beta}\nabla_{\alpha,\beta}\T T\|
  \approx
  0
  \ .
\end{equation}

@@

The parameter $\beta$ can be constrained for a given mean to be
\begin{equation}
  \beta = \frac{\lambda\alpha}{\widehat{\mu}}
  \label{eq:compoundPoisson:beta_restrict}
\end{equation}
where
\begin{equation}
  \widehat{\mu} = \frac{1}{n}\sum_{i=1}^n x_i
\end{equation}
and it was investigated whenever a unique solution to $\dfrac{\partial T}{\partial \alpha} = 0$ can be found for high $\lambda$ and $\alpha$. The constrained objective function is
\begin{multline}
  T(\lambda,\alpha)
  =
  -n\lambda+
  \sum_{i=1}^n
  \mathbb{I}(x_i>0)
  \left[
  y_i\alpha
  \ln\left(\frac{\lambda\alpha}{\widehat{\mu}}\right)-\ln\Gamma(\alpha y_i) - \frac{1}{2}\zeta_i\alpha^2\psi'(y_i\alpha)
  \right.
  \\
  \left.
    +y_i\alpha\ln x_i - \frac{\lambda\alpha x_i}{\widehat{\mu}}
    + y_i \ln \lambda
    \vphantom{\frac{1}{1}}
  \right]
  + c
\end{multline}
which can be approximated to
\begin{align}
  T(\lambda,\alpha)&=
  \begin{multlined}[t]
    -n\lambda+
    \sum_{i=1}^n
    \mathbb{I}(x_i>0)
    \left[
      y_i\alpha\ln\left(\frac{\lambda\alpha}{\widehat{\mu}}\right)-\alpha y_i\ln(\alpha y_i) + \alpha y_i - \frac{1}{2}\frac{\zeta_i\alpha}{y_i}
    \right.
    \\
    \left.
      +y_i\alpha\ln x_i - \frac{\lambda\alpha x_i}{\widehat{\mu}}
      + y_i \ln \lambda
      \vphantom{\frac{1}{1}}
    \right]
    + c
  \end{multlined}
  \nonumber\\
  &\!\begin{multlined}[b]=
    -n\lambda+
    \sum_{i=1}^n
    \mathbb{I}(x_i>0)
    \left[
      y_i\alpha\ln\left(\frac{\lambda}{y_i\widehat{\mu}}\right) + \alpha y_i - \frac{1}{2}\frac{\zeta_i\alpha}{y_i}
    \right.
    {}\\
    \left.
      +y_i\alpha\ln x_i - \frac{\lambda\alpha x_i}{\widehat{\mu}}
      + y_i \ln \lambda
      \vphantom{\frac{1}{1}}
    \right]
    + c
    \ .
  \end{multlined}
\end{align}
Taking the derivative with respect to $\alpha$
\begin{equation}
  \frac{\partial T}{\partial \alpha} =
  \sum_{i=1}^n
  \mathbb{I}(x_i>0)
  \left[
    y_i\ln\left(\frac{\lambda}{y_i\widehat{\mu}}\right)
    + y_i - \frac{1}{2}\frac{\zeta_i}{y_i}
    +y_i\ln x_i - \frac{\lambda x_i}{\widehat{\mu}}
    \vphantom{\frac{1}{1}}
  \right]
\end{equation}
and this is not $\alpha$ dependent, thus there is no solution to $\alpha$.

@@

The log likelihood, using a simulation of 100 compound Poisson-gamma random variables, are plotted for fixed $\lambda$ in Figure \ref{fig:compoundPoisson_loglikelihoodplot}. These plots are concerning because the likelihood appeared not to be convex and there is no unique maxima for all $\lambda$ investigated.

@@

As a results, there is no unique maximum likelihood estimator and the EM algorithm would be very sensitive to the starting position, including for small $\lambda$.

@@

XCT is in the realm of high $\lambda$ and this can be shown using back of the envelope calculations. Using the properties in the \texttt{AbsNoFilter} dataset, the power and potential difference of the x-ray tube are $P=\SI{36}{\watt}$ and $V=\SI{80}{\kilo\volt}$ respectively. The time exposure is $\tau=\SI{708}{\milli\second}$. Assume an efficiency of \SI{0.01}{{photon.}{electron}^{-1}} and the charge per electron is $e=\SI{1.602e-19}{\coulomb\,{electron}^{-1}}$. Assume the photons projected onto the array of $2\,000^2\,\si{\pixel}$. The number of photons detected per pixel is
\begin{align*}
\lambda &\approx \dfrac{\text{efficiency} \times P \tau}{\text{number of pixels} \times V e}
\\
&\sim 10^7\,\si{{photon}\,\pixel^{-1}}
\ .
\end{align*}
Fitting the compound Poisson model onto XCT data would not be possible with such a high value of $\lambda$.
